---
title: "Reproducible Research - Course Project 1"
date: "26 augusti 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
```

## Introduction
This document will answer the questions asked in the assignment instructions. The document is structured into the questions asked.

## Load required packages
```{r, message = FALSE, warning = FALSE}
library(ggplot2)
library(dplyr)
library(data.table)
library(gridExtra)
```

## Loading and preprocessing the data
```{r}
# Download the file
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", destfile = "activity.zip")

# Unzip the file
unzip("activity.zip")

# Make a data frame containing the data in the .csv file
activityDF <- read.csv("activity.csv", colClasses = c("numeric", "Date", "integer"))
```

## What is mean total number of steps taken per day?
```{r}
# Calculate the total amount of steps per day
activityDF2 <- activityDF %>% group_by(date) %>% summarize(steps_per_day = sum(steps, na.rm = TRUE))

# Make a histogram showing the distribution of steps per day
ggplot(activityDF2, aes(steps_per_day)) + geom_histogram(bins = 20) + xlab("steps per day") + ggtitle("Counts of number of steps taken per day")

# Calculate the mean of total number of steps taken per day
mean(activityDF2$steps_per_day, na.rm = TRUE)

# Calculate the median of total number of steps taken per day
median(activityDF2$steps_per_day, na.rm = TRUE)

```

## What is the average daily activity pattern?
```{r}
# Calculate the mean number of steps per interval
activityDF3 <- activityDF %>% group_by(interval) %>% summarize(mean_steps = mean(steps, na.rm = TRUE))

# Plot the diagram
ggplot(activityDF3, aes(interval, mean_steps)) + geom_line() + xlab("time interval") + ylab("mean steps per 5-minute interval")

# The interval with the highest average of steps
activityDF3[which.max(activityDF3$mean_steps), ]
```

## Imputing missing values
```{r}
# Calculate the number of rows with missing values
sum(!complete.cases(activityDF))
```

The NA values will be replaced by the mean number of steps for the corresponding interval.
To more easily impute these values, we will make data.table copies of activityDF and activityDF3,
since activityDF3 holds the interval means.

```{r}
# Create data.table copies of activityDF and activityDF3
activityDT <- as.data.table(activityDF)
activityDT3 <- as.data.table(activityDF3)

# Set the interval column to be a key in each data.table
setkey(activityDT, interval)
setkey(activityDT3, interval)

# Join the two tables and create a new column in activityDT holding the values to be imputed
activityDT[activityDT3, imputed_steps := i.mean_steps]

# Replace the NAs with the imputed values
activityDT[is.na(steps), steps := imputed_steps]

# Delete the column containing the temporary imputed values.
activityDT <- activityDT[, 1:3]

# Calculate the total amount of steps per day
activityDT2 <- activityDT %>% group_by(date) %>% summarize(steps_per_day = sum(steps, na.rm = TRUE))

# Make a histogram showing the distribution of steps per day
ggplot(activityDT2, aes(steps_per_day)) + geom_histogram(bins = 20) + xlab("steps per day") + ggtitle("Counts of number of steps taken per day")

# Calculate the mean of total number of steps taken per day
mean(activityDT2$steps_per_day, na.rm = TRUE)

# Calculate the median of total number of steps taken per day
median(activityDT2$steps_per_day, na.rm = TRUE)
```

Imputing values, based on the mean number of steps of the corresponding interval, increased the both the mean and median values.

## Are there differences in activity patterns between weekdays and weekends?
```{r}
# Add a column specifying the weekday of the date
activityDT[, weekday := weekdays(date, abbreviate = TRUE)]

# Add a column telling whether the weekday is in the week or weekend
activityDT[, weekday_type := ifelse(weekday == "lör" | weekday == "sön", "weekend", "weekday")]

# Delete the weekday column
activityDT <- activityDT[, c(1,2,3,5)]

# Convert the weekday_type column to a factor variable
activityDT$weekday_type <- as.factor(activityDT$weekday_type)

# Calculate the mean number of steps per interval
activityDT4 <- activityDT %>% group_by(interval, weekday_type) %>% summarize(mean_steps = mean(steps))

# Create a ggplot object for the weekend part
g1 <- ggplot(activityDT4[activityDT4$weekday_type == "weekend",], aes(interval, mean_steps)) + geom_line(color = "blue") + xlab("time interval") + ylab("mean steps per 5-minute interval") + ggtitle("weekend")

# Create a gglot object for the weekday part
g2 <- ggplot(activityDT4[activityDT4$weekday_type == "weekday",], aes(interval, mean_steps)) + geom_line(color = "red") + xlab("time interval") + ylab("mean steps per 5-minute interval") + ggtitle("weekday")

# Plot the two diagrams at the same time
grid.arrange(g1, g2, nrow = 2)
```
